---
title: Enabling Data Science in DIPS 
date: 2020-04-23
draft: false
author: Bjørn Fjukstad
---

At DIPS we are currently productionizing our new electronic health record (EHR)
system, DIPS Arena. With DIPS Arena we have made the move to store structured
clinical data with [openEHR](https://openehr.org). openEHR is an open technology
for e-health systems that help different vendors to build interoperable systems
for healthcare. In other words, it helps define the different clinical concepts,
e.g.  blood pressures, body temperatures etc, how we can represent them, and how
we can retrieve them from systems that implement an openEHR API. While there is
a lot of detailed information on openEHR, I believe that we only have to care
about how we can get data from an openEHR server to begin analyzing data in our
EHR system. If you want to learn more about openEHR, please take
a tour of [openehr.org](https://openehr.org).

# How to get data 
To get data from DIPS Arena, or an openEHR server, we execute [Archetype Query
Language (AQL)](https://specifications.openehr.org/releases/QUERY/latest/AQL.html) 
queries. These look like your usual database queries, with
`select`, `from`, and `where` clauses. For example if we want to get all blood
pressures stored in our EHR system we can execute the query 


```sql
	select c
	from composition c
	contains observation o[openEHR-EHR-OBSERVATION.blood_pressure.v1]
```

which will return either an `XML` or `JSON` representation of the result.

To summarize the above query, we select all compositions that contain an blood
pressure (`openEHR-EHR-OBSERVATION.blood_pressure.v1`) observation. A
Composition is a container where we store information. All information stored
within an openEHR system is contained within compositions. These compositions
may hold any type of data, so by supplying a `contains` we can specify that we
only want the compositions with a specific `observation`. Observations are
special containers that are ment for recording clinical observations and any
additional metadata. Typical observations are blood pressures, body weights,
heights. Metadata may be information about the data collection method, what type
of instrument was used, etc. 


The first lines from the JSON result from the above query are as follows: 

```json
{
  "_type": "RESULTSET",
  "meta": {
    "_type": "RESTULTSET",
    "_schema_version": "1.0.0",
    "_created": "2020-04-14T08:28:38.1538856Z",
    "_generator": "DIPS.openEHR.dll (5.0.0.0)"
  },
  "name": null,
  "totalResults": 427,
  "columns": [
    {
      "name": "c",
      "path": ""
    }
  ],
  "rows": [
    [
      {
        "_type": "COMPOSITION",
        "archetype_node_id": "openEHR-EHR-COMPOSITION.encounter.v1",
        "name": {
          "value": "Klinisk kontakt"
        },
        "uid": {
          "_type": "OBJECT_VERSION_ID",
          "value": "7da88898-0a36-4ca3-b972-111999ed16ee::dips::1"
        },

```

To summarize, the resultset contains 427 compositions, and these can be found in
the `rows` object of the resultset.

Since we are usually just interested in the actual measurements (systolic and
diastolic) of a blood pressure, we can re-write the AQL query to get just the
measurements we are interested in: 


```sql
select
    c/content[openEHR-EHR-OBSERVATION.blood_pressure.v1]/data[at0001]/events[at0006]/time/value as datetime,
    c/content[openEHR-EHR-OBSERVATION.blood_pressure.v1]/data[at0001]/events[at0006]/data[at0003]/items[at0004]/value/magnitude AS systolic,
    c/content[openEHR-EHR-OBSERVATION.blood_pressure.v1]/data[at0001]/events[at0006]/data[at0003]/items[at0005]/value/magnitude AS diastolic
from
    composition c
    contains observation o[openEHR-EHR-OBSERVATION.blood_pressure.v1]

```

This query is a bit more complex. Especially the [openEHR path syntax](https://specifications.openehr.org/releases/QUERY/latest/AQL.html#_openehr_path_syntax)
(the `c/content[openEHR-EHR ... /time/value` part of the query) might be a bit
unfamiliar. This is the syntax we use to extract specific parts of the stored
information. The openEHR paths in the query will depend on the data we are
querying for, but we can execute the same query on any openEHR implementation
and get the same results. 

In short the query get the `datetime` when a blood pressure was measured, as
well as the `systolic` and `diastolic` measurements of the blood pressure. The
resultset now looks like, where the measurements are found in the `rows` array. 

```json
{
  "_type": "RESULTSET",
  "meta": {
    "_type": "RESTULTSET",
    "_schema_version": "1.0.0",
    "_created": "2020-04-14T10:52:16.3735268Z",
    "_generator": "DIPS.openEHR.dll (5.0.0.0)"
  },
  "name": null,
  "totalResults": 427,
  "columns": [
    {
      "name": "datetime",
      "path": "/content[openehr-ehr-observation.blood_pressure.v1]/data[at0001]/events[at0006]/time/value"
    },
    {
      "name": "systolic",
      "path": "/content[openehr-ehr-observation.blood_pressure.v1]/data[at0001]/events[at0006]/data[at0003]/items[at0004]/value/magnitude"
    },
    {
      "name": "diastolic",
      "path": "/content[openehr-ehr-observation.blood_pressure.v1]/data[at0001]/events[at0006]/data[at0003]/items[at0005]/value/magnitude"
    }
  ],
  "rows": [
    [
      "2010-06-08T10:26:51.037+02:00",
      124.37659794247064,
      77.51167721740681
    ],
    [
      "2011-01-15T19:53:52.844+01:00",
      115.18465581503439,
      80.17813862922695
    ],
```

# Analyzing openEHR data in R 
The [R](https://www.r-project.org) programming language, along with the
[tidyverse](https://www.tidyverse.org) makes it is easy to get started with data
analysis. And to make it simpler for us to get data from an openEHR server, we
have developed the `openehR` R package!

```{r}
library(openehR)
```

```{r, include=F}
library(dplyr)
library(tidyr)
library(ggplot2)
library(kableExtra)
url = "https://vt-selecta-b.dips.local:4443/openehr/"
```

With the `openehR` package we simply specify where the server is running, and
can then begin executing queries.`url` is the address of the openEHR REST API 
endpoint, e.g. `http://our-openehr-server/openehr/`. 

```{r}
query <- "
select
    c/content[openEHR-EHR-OBSERVATION.blood_pressure.v1]/data[at0001]/events[at0006]/time/value as datetime,
    c/content[openEHR-EHR-OBSERVATION.blood_pressure.v1]/data[at0001]/events[at0006]/data[at0003]/items[at0004]/value/magnitude AS systolic,
    c/content[openEHR-EHR-OBSERVATION.blood_pressure.v1]/data[at0001]/events[at0006]/data[at0003]/items[at0005]/value/magnitude AS diastolic
from
    composition c
    contains observation o[openEHR-EHR-OBSERVATION.blood_pressure.v1]
"
    
data =  openehR::query(url, query)

kable(head(data), format="markdown")
```




Since this is an initial version of the `openehR` package, we leave it 
up to the caller to convert the variables to the correct data type. In 
future versions we'll do this automatically. But, with `dplyr` it's 
an easy step to make sure that the dates are actual `date_time`s and
that the blood pressure measurements are numeric. 

```{r}
data = data %>% 
  mutate(datetime = lubridate::parse_date_time(datetime, "YmdHMOSz"))  %>% 
  mutate(systolic = as.numeric(as.character(systolic))) %>% 
  mutate(diastolic = as.numeric(as.character(diastolic)))  %>% 
  arrange(desc(datetime))
```

With the variables converted we can start plotting the data. Let's have a 
look at the measurements between 2010 and 2018. 


```{r}
data  %>% 
  filter(datetime > lubridate::ymd("2010-01-01")) %>% 
  filter(datetime < lubridate::ymd("2018-01-01")) %>% 
  ggplot(aes(x=datetime)) + 
    geom_line(aes(y=systolic, color="systolic")) + 
    geom_line(aes(y=diastolic, color="diastolic")) + 
  theme_classic() + 
  theme(plot.title.position = "plot") +
  labs(title="Blood Pressures collected between 2010 and 2018", 
       color="Measurement") + 
  ylab("mmHg") + 
  xlab("Date") + 
  ylim(0,200)
  
```

We could go on and on with more visualizations, and analysis of the different
data stored in our openEHR server, but I think this will do for this blog post.
I hope I have given you an example of how we can leverage the structured
clinical data stored in an openEHR server, and specifically DIPS Arena. 

If you have any questions or comments, please drop me an e-mail at
[bfj@dips.no](mailto:bfj@dips.no)!  
